---
title: "Random Matrix Theory"
author: "Evgenii Chzhen"
date: "16 Jan 2016"
output:
  pdf_document:
    fig_crop: no
    fig_height: 3.5
    toc: yes
  html_document:
    theme: cosmo
    toc: yes
header-includes:
- \usepackage{graphicx}
- \usepackage{bbm}
- \usepackage{mathtools}
- \usepackage{amsthm}
- \usepackage{amssymb}
- \usepackage{amsmath}
- \newtheorem{mydef}{Definition}
- \newtheorem{myrem}{Remark}
- \newtheorem{thm}{Theorem}
bibliography: bibliography.bib
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library("mvtnorm",quietly=TRUE)
library("cmvnorm",quietly=TRUE)
library("RMTstat",quietly=TRUE)

# Form a Wigner matrix. Sigma is a variance of X_{i,j}. 
# There're 2 real valued distributions (normal, uniformal) and one complex valued distribution (normal)
simulateWignerMatrix <- function(n, sigma=1, distr='normal'){
  if(distr=='normal'){
    X <- matrix(rnorm(n*n, 0, sigma), n, n)
    X <- X / sqrt(n)
    X[lower.tri(X)] = t(X)[lower.tri(X)]
    return(X)
    } 
  else if(distr=='uniform'){
    X <- matrix(runif(n*n, min=-sqrt(3)*sigma, max=sqrt(3)*sigma), n, n)
    X <- X / sqrt(n)
    X[lower.tri(X)] = t(X)[lower.tri(X)]
    return(X)
    }
  else if(distr=='complex_normal'){
    X <- matrix(rcmvnorm(n*n, mean=c(0), sigma=matrix(sigma^2, 1, 1)), n, n)
    X <- X / sqrt(n)
    X[lower.tri(X)] = Conj(t(X))[lower.tri(X)]
    return(X)
    }
}

# Form a Marchenko-Pastur matrix. Sigma is a variance of X_{i,j}. 
# There're 2 real valued distributions (normal, uniformal) and one complex valued distribution (normal)
simulateMarchenkoPasturMatrix <- function(N, n, sigma=1, distr='normal'){
  if(distr=='normal'){
    X <- matrix(rnorm(N*n, 0, sigma), N, n)
    return(X)
    } 
  else if(distr=='uniform'){
    X <- matrix(runif(N*n, min=-sqrt(3)*sigma, max=sqrt(3)*sigma), N, n)
    return(X)
    }
  else if(distr=='complex_normal'){
    X <- matrix(rcmvnorm(N*n, mean=c(0), sigma=matrix(sigma^2, 1, 1)), N, n)
    return(X)
    }
}

# Computes eigenvalues of given Matrix and plots histogram.
# if freal=TRUE also plots a semi-circle distributionhttp://127.0.0.1:27270/graphics/plot_zoom_png?width=956&height=480
drawWignerHistogram <- function(X, sigma=1, freal=FALSE){
  semiCircle <- function(x) sqrt(4 * sigma^2 - x^2) / (2 * pi * sigma^2)
  eigenValues <- eigen(X, symmetric=TRUE ,only.values=TRUE)$values
  a = paste("N = ", toString(dim(X)[1]))
  hist(eigenValues, breaks=35, freq=FALSE, main=a , xlab='Eigen Values', ylab='Density' )
  if(freal) curve(semiCircle, add=TRUE, col='red')
}

drawMarchenkoPasturHistogram <- function(X, sigma=1, freal=FALSE){
  c = dim(X)[1] / dim(X)[2]
  Z = (1 / dim(X)[2]) * (X %*% Conj(t(X)))
  eigenValues <- eigen(Z, symmetric=TRUE, only.values=TRUE)$values
  a = paste("N =", toString(dim(X)[1])) 
  b = paste(" n =", toString(dim(X)[2]))
  a = paste(a, b)
  hist(eigenValues, breaks=35, freq=FALSE, main=a, xlab='Eigen Values', ylab='Density' )
  if(freal) curve(dmp(x, ndf=NA, pdim=NA, var=sigma, svr=1/c, log = FALSE), n=1e4, add=TRUE, col='red')
}

```

This paper is based on lectures given by Mr. Jamal Najim at Universit?? Paris-Est Marne-la-Vall??e.
## 1. Introduction
In this paper we want to proceed experimental tests of two theorems in random matrix theory (RMT). We start with an introduction of basic objects in RMT. 

\begin{mydef}
Consider a symmetric (Hermitian) matrix $Z_{N}$ of $N\times N$ dimension, each element of the matrix $(Z_N)_{ij}$ is a random real (complex) valued variable. An empirical spectral measure of $Z_{N}$ is defined as
$$ \mathcal{L}_N(A) = \frac{1}{N}\sum\limits_{i=1}^N\delta_{\lambda_i}(A) = \frac{\#\{\textrm{eigenvalues of }Z_N\textrm{ in }A\}}{N}, \textrm{    for } A \subseteq \mathbb{R}. $$
\end{mydef}

\begin{myrem}
Since $Z_{N}$ is a symmetric (Hermitian) matrix, therefore by Spectral theorem there exist $\lambda_1, ..., \lambda_N \in \mathbb{R}$, where $\lambda_1, ..., \lambda_N$ are eigenvalues of matrix $Z_N$ counted with multiplicity.
\end{myrem}

\begin{myrem}
$\forall f \in \mathbb{C}_b(\mathbb{R})$ holds $\int f(x)\,\mathcal{L}_N(dx) = \frac{1}{N}\sum\limits_{i=1}^Nf(\lambda_i).$
\end{myrem}

## 2. Wigner theorem

In this section we define a Wigner matrix and state a Wigner theorem. 

\begin{mydef}
$Z_N$ is called a Wigner Matrix if $Z_N$ is a Hermitian matrix such that for all $i \leq j$ $(Z_N)_{ij} = \frac{Y_{ij}}{\sqrt{N}}$ where $Y_{ij}$ are i.i.d random variables with $\mathbb{E}Y_{ij} = 0$ and $VarY_{ij} = \sigma^2$.
\end{mydef}

\begin{mydef}
 Random variable $X$ has a semicircle distribution if its' probability density function is defined as it follows

$$ \mathbb{P}_{sc}(dx) = \frac{\sqrt{4\sigma^2 - x^2}}{2 \pi\sigma^2}\mathbbm{1}_{[-2\sigma,2\sigma]}(x)dx. $$
\end{mydef}

\begin{thm}
Let $Z_N$ be a Wigner Matrix and $\mathcal{L}_N$ is an empirical spectral measure of $Z_N$, therefore 
almost surely $$ \mathcal{L}_N \xrightarrow{\mathcal{D}}\mathbb{P}_{sc} $$.
\end{thm}

Now we want to make a numerical experiments, let $Z_N$ be a Wigner Matrix, for $N = 50, 300 ,1000$ we make a simulation of the matrix from a given distribution and compute eigenvalues after we plot a histogram of obtained values and compare them with theoretical results. Red line corresponds to semicircle distribution.

```{r, echo=FALSE, fig.height=2.5, warning=FALSE, message=FALSE}
par(mfrow = c(1, 3), oma=c(0,0,2,0))
drawWignerHistogram(simulateWignerMatrix(50, 1, 'normal'), 1, freal=FALSE)
drawWignerHistogram(simulateWignerMatrix(300, 1, 'normal'), 1, freal=FALSE)
drawWignerHistogram(simulateWignerMatrix(1000, 1, 'normal'), 1, freal=TRUE)
mtext('Standart Gaussian distribution', outer=TRUE, cex = 1)

par(mfrow = c(1, 3), oma=c(0,0,2,0))
drawWignerHistogram(simulateWignerMatrix(50, 1, 'uniform'), 1, freal=FALSE)
drawWignerHistogram(simulateWignerMatrix(300, 1, 'uniform'), 1, freal=FALSE)
drawWignerHistogram(simulateWignerMatrix(1000, 1, 'uniform'), 1, freal=TRUE)
mtext('Uniform distribution', outer=TRUE, cex = 1)

par(mfrow = c(1, 3), oma=c(0,0,2,0))
drawWignerHistogram(simulateWignerMatrix(50, 1, 'complex_normal'), 1, freal=FALSE)
drawWignerHistogram(simulateWignerMatrix(300, 1, 'complex_normal'), 1, freal=FALSE)
drawWignerHistogram(simulateWignerMatrix(1000, 1, 'complex_normal'), 1, freal=TRUE)
mtext('Complex Gaussian distribution', outer=TRUE, cex = 1)
```


## 3. Marchenko-Pastur theorem

In this section we state a Marchenko-Pastur theorem.

\begin{mydef}
For $c > 0$ a distribution of Marchenko-Pastur is defined as
$$\mathbb{P}_{MP}(dx) = 
\begin{cases} \frac{\sqrt{(\lambda^+-x)(x-\lambda^-)}}{2\pi x\sigma^2c}\mathbbm{1}_{[\lambda^-,\lambda^+]}(x)dx, & \textrm{if } 0 < c \leq 1 \\ 
(1-\frac{1}{c})\delta_0(dx) + \frac{\sqrt{(\lambda^+-x)(x-\lambda^-)}}{2\pi x\sigma^2c}\mathbbm{1}_{[\lambda^-,\lambda^+]}(x)dx, & \textrm{if } c>1 \end{cases},$$
where $\lambda^+ = \sigma^2(1+\sqrt{c})^2$ and $\lambda^- = \sigma^2(1-\sqrt{c})^2$. 
\end{mydef}

\begin{thm}
Let $Z_N = \frac{1}{n}X_NX_N^*$, where $X_N$ is a matrix of dimension $N\times n$ and $(X_{ij})_{ij}$ are i.i.d. random variables such that $\mathbb{E}X_{ij} = 0$ and $Var X_{ij} = \sigma^2$. Let $\mathcal{L}_N$ be an empirical spectral measure of $Z_N$. If $\frac{N}{n} \rightarrow c > 0$ therefore almost surely
$$ \mathcal{L}_N \xrightarrow{\mathcal{D}}\mathbb{P}_{MC}, $$
where $\mathbb{P}_{MC}$ is a Marchenko-Pastur distribution with parameters $c, \sigma^2$.
\end{thm}

\begin{myrem}
Assume that $c > 1$ then for large enough $N, n$ we have $N > n$ since $rank(Z_N) = min\{N, n\}$ we will have $N - n = n(c - 1)$ zero eigenvalues.
Since $\mathcal{L}_N = \frac{1}{N}\sum\limits_{i=1}^N\delta_{\lambda_i}$ we can observe that in the limiting measure there will be a mass of $\frac{n(c - 1)}{N} = (1-\frac{1}{c})$ at $0$ point.
\end{myrem}

Now we want to make a numerical experiments, let $Z_N$ be a matrix which satisfies assumptions of Marchenko-Pastur theorem, we make a simulation of the matrix from a given distribution and compute eigenvalues after we plot a histogram of obtained values and compare them with theoretical results. Red line corresponds to Marchenko-Pastur distribution.

```{r, echo=FALSE, cache=TRUE}
par(mfrow = c(1, 3), oma=c(0,0,2,0))
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(50, 500, distr='normal'), freal=FALSE)
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(250, 2500, distr='normal'), freal=FALSE)
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(1000, 10000, distr='normal'), freal=TRUE)
mtext('Standart Gaussian distribution (c = 0.1)', outer=TRUE, cex = 1)

par(mfrow = c(1, 3), oma=c(0,0,2,0))
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(40, 40, distr='uniform'), freal=FALSE)
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(400, 400, distr='uniform'), freal=FALSE)
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(1000, 1000, distr='uniform'), freal=TRUE)
mtext('Uniform distribution (c = 1)', outer=TRUE, cex = 1)

par(mfrow = c(1, 3), oma=c(0,0,2,0))
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(150, 125, distr='complex_normal'), freal=FALSE)
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(600, 500, distr='complex_normal'), freal=FALSE)
drawMarchenkoPasturHistogram(simulateMarchenkoPasturMatrix(500, 10000, distr='complex_normal'), freal=TRUE)
mtext('Complex Gaussian distribution (c = 1.2)', outer=TRUE, cex = 1)
```

## 4. Corollary

In this paper we introduced some basic definitions and theorems of random matrix theory, we proceeded a numerical procedure to show the ideas of theorems on simulated data. Our experimental results are in high agreement with the theoretical. For numerical experiments we used R language, particulary **RMTstat** package were used for a Marchenko-Pastur distribution and **cmvnorm** package for simulating complex valued Gaussian variables.

